---
title: "Work | Michael Esbach"
project:
  type: website

format:
  html:
    page-layout: full
    link-external-newwindow: true

execute:
  echo: false

lightbox: true
---

<div class="work-grid">

<div class="work-box">
<div class="work-header">
<h2>Territorial Resilience</h2>
</div>
<div class="work-content">
<div class="work-image">
```{r, echo=FALSE, message=FALSE, warning=FALSE}
source("carousel/carousel-work.R")
carousel("carousel-work1", duration = 10000, items = list(
  list(caption = "Siekopai", image = "images/work/web/territory1.jpg"),
  list(caption = "Finishing up a transect", image = "images/work/web/territory2.jpg"),
  list(caption = "Siona", image = "images/work/web/territory3.jpg"),
  list(caption = "Truck ride", image = "images/work/web/territory4.jpg"),
  list(caption = "Siekopai territory", image = "images/work/web/territory5.jpg")
))
```
</div>
<div class="work-text">
Much of my current work explores how Indigenous peoples collectively sustain biocultural processes and territorial resilience of the Amazon. As a Co-Principal Investigator on a National Science Foundation Dynamics of Integrated Socio-Environmental Systems (DISES) [grant](https://www.nsf.gov/awardsearch/showAward?AWD_ID=2108308&HistoricalAwards=false), I collaborate with Cofán, Siekopai, and Siona communities to investigate biocultural stewardship strategies and their impact on territorial resilience. Through transdisciplinary methods, I lead capacity-building initiatives, co-develop culturally grounded research tools, and support Indigenous researchers in analyzing and applying findings to enhance self-determined stewardship.

-   Correia, J. E., Piaguage Lucitante, J., Weiss, L., Narváez, N., Lucitante, L., Biaguaje, A., ... & **Esbach, M.** (2025). Stimulating Reciprocity: How Human–plant Relations Support Indigenous Cultural Revitalization and Stewardship in the Ecuadorian Amazon. *People and Nature*, 7, 1151–1170. \[[doi](https://doi.org/10.1002/pan3.70019), [Relational Thinking Blog](https://relationalthinkingblog.com/2025/02/18/plain-language-summary-stimulating-reciprocity-how-human-plant-relations-support-indigenous-cultural-revitalization-and-stewardship-in-the-ecuadorian-amazon/)\]

-   Correia, J. E., Walker, R. T., Simmons, C. S., Urgilés, C., Narváez Quenama, N., Yiyocuro, L., ... & **Esbach, M.** (2025). Biocultural Geographies: Stewardship, Indigenous Territories, and Conservation in Ecuador's Amazon. *Annals of the American Association of Geographers*, 1-19. \[[doi](https://doi.org/10.1080/24694452.2025.2511942)\]
</div>
</div>
</div>

<div class="work-box">
<div class="work-header">
<h2>Conservation and Care</h2>
</div>
<div class="work-content">
<div class="work-image">
```{r}
source("carousel/carousel-work.R")
carousel("carousel-work2", duration = 10000, items = list(
  list(caption = "Cooling off in the Zábalo River", image = "images/work/web/cofan1.jpg"),
  list(caption = "Community meeting", image = "images/work/web/cofan2.jpg"),
  list(caption = "Looking for timber", image = "images/work/web/cofan3.jpg"),
  list(caption = "Setting camera traps", image = "images/work/web/cofan4.jpg"),
  list(caption = "Prepping fishing net", image = "images/work/web/cofan5.jpg")
))
```
</div>
<div class="work-text">
This area of work focuses on how Indigenous practices of care, reciprocity, and governance sustain ecological abundance, cultural vitality, and territorial integrity in Amazonia. Through long-term collaboration with Cofán communities and regional analyses of Indigenous territories, I investigate how Indigenous stewardship contributes to biodiversity conservation and more equitable conservation models.

-   **Esbach, M.**, Urgilés-Verdugo, C., Townsend, W., & Yiyoguaje, C. (2024). Hunting for Sustainability in the Cofán Territory of Zábalo, Ecuadorian Amazon. *Conservation Letters*, 17(6), e13065. \[[doi](https://doi.org/10.1111/conl.13065)\]

-   **Esbach, M.**, Correia, J. E., Valdivia, G., & Lu, F. (2024). Amazonian Conservation across Archipelagos of Indigenous Territories. *Conservation Biology*, 39(2), e14407.\[[doi](https://doi.org/10.1111/cobi.14407), [Mongabay](https://es.mongabay.com/2024/12/territorios-indigenas-amazonicos-fragmentados-tambien-son-sitios-importantes-de-conservacion-estudio/)\]

-   **Esbach, M.**, Lu, F., Silva, N., & Borman-Quenama, F. (2024). Conservation and Care: Cofán Lessons for Stewarding Abundance in Amazonia. *Human Ecology*, 52(4), 837–850. \[[doi](https://doi.org/10.1007/s10745-024-00532-2)\]

-   **Esbach, M.**, Puri, M., Botero-Arias, R., & Loiselle, B. A. (2024). Beyond the Island: Integrated Approaches to Conserving Biodiversity Islands with Local Communities. In: *Biodiversity Islands: Pockets of Protected Land in Human Dominated Environments* (pp. 551-568). Cham: Springer International Publishing. \[[doi](https://doi.org/10.1007/978-3-030-92234-4_22)\]
</div>
</div>
</div>

<div class="work-box">
<div class="work-header">
<h2>Ecology and Natural History</h2>
</div>
<div class="work-content">
<div class="work-image">
```{r}
source("carousel/carousel-work.R")
carousel("carousel-work", duration = 10000, items = list(
  list(caption = "Seagrass survey", image = "images/work/web/ecology1.jpg"),
  list(caption = "Hunting peccary", image = "images/work/web/ecology2.jpg"),
  list(caption = "Tapir print", image = "images/work/web/ecology3.jpg")
))
```
</div>
<div class="work-text">
Through collaborative research with Indigenous communities, much of my research seeks to understand the diversity of life — human and more-than-human. Together, we document species distributions, ecological dynamics, and cultural relationships with animals, contributing to both biodiversity science and Indigenous knowledge systems.

-   Urgilés-Verdugo, C., Gallo-Viracocha, F., **Esbach, M.**, & Escudero, A. (2025). Densidad y ocupación del tapir andino en los Andes del Norte del Ecuador. *Ecosistemas*, 34(1), 2859-2859. \[[doi](https://doi.org/10.7818/ECOS.2859)\]

-   Urgilés-Verdugo, C., Payaguaje, L., Biaguaje, A., Payaguaje, H., & **Esbach, M.** (2024). Anomalías en la coloración de piel y pelo de Dicotyles tajacu (Artiodactyla: Tayassuidae) y Dasyprocta fuliginosa (Rodentia: Dasyproctidae) en la Amazonía ecuatoriana. *Mammalogy Notes*, 10(2), 412-412. \[[doi](https://doi.org/10.47603/mano.v10n2.412)\]

-   Payaguaje, H.... **Esbach, M.** (2023). Animales del Territorio Siekopai, Sucumbíos, Ecuador. *Nationalidad Siekopai*. \[[doi](10.6084/m9.figshare.24922029)\]

-   Yiyocuro, H.... **Esbach, M.** (2023). Animales del Territorio Siona, Sucumbíos, Ecuador. *Nationalidad Siona*. \[[doi](10.6084/m9.figshare.24922083)\]

-   Quenama, A.... **Esbach, M.** (2023). Animales del Territorio Cofán de Sinangoe, Sucumbíos, Ecuador. *Comunidad Cofán de Sinangoe*. \[[doi](10.6084/m9.figshare.24922053)\]
</div>
</div>
</div>

<div class="work-box">
<div class="work-header">
<h2>Field Work and Methods</h2>
</div>
<div class="work-content">
<div class="work-image">
```{r}
source("carousel/carousel-work.R")
carousel("carousel-work4", duration = 10000, items = list(
  list(caption = "Transect surveys", image = "images/work/web/methods1.jpg"),
  list(caption = "Camera trap surveys", image = "images/work/web/methods2.jpg"),
  list(caption = "Camera trap surveys", image = "images/work/web/methods3.jpg"),
  list(caption = "Mapping work", image = "images/work/web/methods4.jpg")
))
```
</div>
<div class="work-text">
As an interdisciplinary researcher, I combine ecological monitoring with participatory methods to better understand and support conservation in complex environments. My work includes developing tools for estimating animal populations in dense forests, collaborating with Indigenous trackers, and applying approaches like camera trapping, participatory mapping, and spatial analysis to tackle practical challenges in biodiversity research.

-   **Esbach, M.** (2023). Estimating Mammal Density from Track Counts Collected by Indigenous Amazonian Hunters. *Perspectives in Ecology and Conservation*, 21(3), 247–252. \[[doi](https://doi.org/10.1016/j.pecon.2023.07.005)\]

-   **Esbach, M.** & Patra, R. K. (2022). Distance Sampling from Curving Transects in Dense Tropical Forests. *Biotropica*, 54(5), 1182-1189. \[[doi](https://doi.org/10.1111/btp.13141)\]
</div>
</div>
</div>

</div>

<style>

/* hide title block but get browser tab title*/
.quarto-title-block {
  display: none !important;
}

/* Work Grid Layout */
.work-grid {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 2rem;
  margin: 2rem 0;
}

/* Work Box Styling */
.work-box {
  border: 1px solid #e0e0e0;
  border-radius: 8px;
  padding: 1.5rem;
  background: transparent;
  transition: all 0.3s ease;
  box-shadow: 0 2px 4px rgba(0,0,0,0.05);
}

.work-box:hover {
  box-shadow: 0 4px 12px rgba(0,0,0,0.1);
}

/* Work Header */
.work-header {
  margin-bottom: 1.5rem;
}

/* Work Content Layout */
.work-content {
  display: flex;
  flex-direction: column;
  gap: 1rem;
}

/* Work Image */
.work-image {
  flex: 0 0 auto;
}

/* Work Text */
.work-text {
  flex: 1;
  font-size: 0.9rem;
  line-height: 1.6;
}

.work-text li {
  font-size: 0.85em;
  margin-bottom: 0.5rem;
}

/* Carousel Styling */
.carousel {
  position: relative;
  width: 100%;
  aspect-ratio: 4/3; /* 4:3 aspect ratio */
  overflow: hidden;
  border-radius: 6px;
  padding-bottom: 3.5rem;
}

.carousel-inner {
  position: relative;
  width: 100%;
  height: 100%;
  overflow: hidden;
}

.carousel-item {
  position: relative;
  width: 100%;
  height: 100%;
}

.carousel-inner img {
  width: 100%;
  height: 100%;
  object-fit: cover;
  display: block;
  border-radius: 6px;
}

.carousel-indicators {
  display: flex;
  justify-content: center;
  align-items: center;
  gap: 8px;
  flex-wrap: nowrap;
}

.carousel-indicators [data-bs-target]:not(.carousel-control-prev):not(.carousel-control-next) {
  width: 8px;
  height: 8px;
  border-radius: 50%;
  background-color: #ccc;
  border: none;
  opacity: 0.5;
  transition: opacity 0.3s ease;
}

.carousel-indicators .active:not(.carousel-control-prev):not(.carousel-control-next) {
  background-color: #666;
  opacity: 1;
}

/* Carousel Control Arrows */
.carousel-control-prev,
.carousel-control-next {
  position: relative;
  top: auto;
  transform: none;
  width: auto;
  height: auto;
  background: none;
  border: none;
  display: flex;
  align-items: center;
  justify-content: center;
  color: #333;
  font-size: 18px;
  text-decoration: none;
  transition: color 0.3s ease;
  z-index: 10;
  cursor: pointer;
  flex-shrink: 0;
  padding: 5px;
  margin: 0;
}

.carousel-control-prev {
  left: auto;
}

.carousel-control-next {
  right: auto;
}

.carousel-control-prev:hover,
.carousel-control-next:hover {
  color: #555;
  text-decoration: none;
}

.carousel-control-prev-icon,
.carousel-control-next-icon {
  width: 20px;
  height: 20px;
  background-size: 20px 20px;
}

.carousel-control-prev-icon {
  background-image: url("data:image/svg+xml,%3csvg xmlns='http://www.w3.org/2000/svg' viewBox='0 0 16 16' fill='%23333'%3e%3cpath d='M11.354 1.646a.5.5 0 0 1 0 .708L5.707 8l5.647 5.646a.5.5 0 0 1-.708.708l-6-6a.5.5 0 0 1 0-.708l6-6a.5.5 0 0 1 .708 0z'/%3e%3c/svg%3e");
}

.carousel-control-next-icon {
  background-image: url("data:image/svg+xml,%3csvg xmlns='http://www.w3.org/2000/svg' viewBox='0 0 16 16' fill='%23333'%3e%3cpath d='M4.646 1.646a.5.5 0 0 1 .708 0l6 6a.5.5 0 0 1 0 .708l-6 6a.5.5 0 0 1-.708-.708L10.293 8 4.646 2.354a.5.5 0 0 1 0-.708z'/%3e%3c/svg%3e");
}

.carousel-caption {
  position: static;
  padding-top: 0.75rem;
  color: #555;
  text-align: center;
  font-size: 0.9rem;
}

/* Mobile Responsive */
@media (max-width: 768px) {
  .work-grid {
    grid-template-columns: 1fr;
    gap: 1.5rem;
    margin: 1rem 0;
  }
  
  .work-box {
    padding: 1rem;
  }
  
  .work-header h2 {
    font-size: 1.2rem;
  }
  
  .work-content {
    gap: 1rem;
  }
  
  .carousel-inner img {
    max-width: 100%;
    height: auto;
  }
}

/* Tablet Responsive */
@media (max-width: 1024px) and (min-width: 769px) {
  .work-grid {
    gap: 1.5rem;
  }
  
  .work-box {
    padding: 1.25rem;
  }
}

</style>

<script>
// Initialize carousel controls
document.addEventListener('DOMContentLoaded', function() {
  // Initialize all carousels
  const carousels = document.querySelectorAll('.carousel');
  
  carousels.forEach(function(carousel) {
    // Initialize Bootstrap carousel if not already initialized
    if (typeof bootstrap !== 'undefined') {
      new bootstrap.Carousel(carousel);
    } else {
      // Fallback for Bootstrap 4
      $(carousel).carousel();
    }
  });
  
  // Add click handlers for carousel controls
  document.querySelectorAll('.carousel-control-prev').forEach(function(control) {
    control.addEventListener('click', function(e) {
      e.preventDefault();
      const target = this.getAttribute('data-bs-target');
      if (target) {
        const carousel = document.querySelector(target);
        if (carousel) {
          if (typeof bootstrap !== 'undefined') {
            const bsCarousel = bootstrap.Carousel.getInstance(carousel);
            if (bsCarousel) {
              bsCarousel.prev();
            } else {
              new bootstrap.Carousel(carousel).prev();
            }
          }
        }
      }
    });
  });
  
  document.querySelectorAll('.carousel-control-next').forEach(function(control) {
    control.addEventListener('click', function(e) {
      e.preventDefault();
      const target = this.getAttribute('data-bs-target');
      if (target) {
        const carousel = document.querySelector(target);
        if (carousel) {
          if (typeof bootstrap !== 'undefined') {
            const bsCarousel = bootstrap.Carousel.getInstance(carousel);
            if (bsCarousel) {
              bsCarousel.next();
            } else {
              new bootstrap.Carousel(carousel).next();
            }
          }
        }
      }
    });
  });
});
</script>
